---
title: "Analysis of Training Data & Simulation on Testing Data"
author: "Nick Kurtansky"
date: "11/20/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(lmtest)
library(ggplot2)
library(dplyr)
library(tidyr)
```


### load in data
```{r}
data <- load('data/favorite.data.RData')
data <- favorite.data
str(data)
```


# begin analysis

### Clean Data
```{r}
# only include matches that sports books took bets on and don't include Round Robins
data <- data %>% filter((FcountOdds > 0) & (UcountOdds > 0) & Round != "Round Robin")

# new column for absolute rank difference variable: abs.rank.dif
data <- data %>% mutate(rank.dif = URank - FRank)

# new column for absolute odds difference variable: abs.odds.dif
data <- data %>% mutate(odds.dif = avgF - avgU)

str(data)
```

### What is the raw probability of an upset?
```{r}
sum(data$odds.upset/nrow(data))
```
Upset occurs with probability .30... Hence, odds have a success rate p = .70

### Training data from 2001 through 2014
```{r}
train <- data %>% filter(!(year %in% c("2015","2016","2017")))
test <- data %>% filter(year %in% c("2015","2016","2017"))
```

### Frequency Tables
```{r}
# Row-proportions: Best.of vs. odds.upset
prop.table(table(train$Best.of, train$odds.upset),1)

# Row-proportions: Surface vs. odds.upset
prop.table(table(train$Surface, train$odds.upset),1)

# Row-proportions: Round vs. odds.upset
prop.table(table(train$Round, train$odds.upset),1)

# Row-proportions: Series vs. odds.upset
prop.table(table(train$Series,train$odds.upset),1)
```

### Frequency Bar Graphs
```{r}
my.theme <- theme(axis.text.x=element_text(face="bold",color="cyan4", size=8, angle=30), axis.text.y=element_text(face="bold",color="cyan4",size=8), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5))

# Frequency Bar Graph: Best.of vs. odds.upset
ggplot(data=train, aes(Best.of,as.numeric(odds.upset))) + stat_summary(fun.y=mean, geom="bar",fill="orange3",width=.4) + labs(y="Frequency (%)", x="Best Of (x) Match", title="Upset Likelihood") + theme(axis.text.x=element_text(face="bold",color="cyan4", size=10), axis.text.y=element_text(face="bold",color="cyan4",size=10), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5))

# Frequency Bar Graph: Surface vs. odds.upset
ggplot(data=train, aes(Surface,as.numeric(odds.upset))) + stat_summary(fun.y=mean, geom="bar",fill="orange3",width=.4) + labs(y="Frequency (%)", x="Court Surface Type", title="Upset Likelihood") + theme(axis.text.x=element_text(face="bold",color="cyan4", size=10), axis.text.y=element_text(face="bold",color="cyan4",size=10), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5))

# Frequency Bar Graph: Round vs. odds.upset
ggplot(data=train, aes(Round,as.numeric(odds.upset))) + stat_summary(fun.y=mean, geom="bar",fill="orange3",width=.4) + labs(y="Frequency (%)", x="Round of Tournament", title="Upset Likelihood") + my.theme

# Frequency Bar Graph: Series vs. odds.upset
ggplot(data=train, aes(Series,as.numeric(odds.upset))) + stat_summary(fun.y=mean, geom="bar",fill="orange3",width=.4) + labs(y="Frequency (%)", x="Series of Tournament", title="Upset Likelihood") + my.theme
```


### Boxplot graphs
```{r}
my.theme <- theme(axis.text.x=element_text(face="bold",color="cyan4", size=8), axis.text.y=element_text(face="bold",color="cyan4",size=8), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5))

# Rank Difference
ggplot(data = train, aes(x = odds.upset, y = rank.dif)) + geom_boxplot(fill = "orange3") + labs(y = "Difference in Player Ranks", x = "Upset (Odds) Occured", title = "Upset Result by Difference in Rank Matchup") + ylim(1,75) + facet_wrap(~Round) + my.theme

# Odds of the Favorite
ggplot(data = train, aes(x = odds.upset, y = FMaxOdds)) + geom_boxplot(fill = "orange3") + labs(y = "Odds of Favorite", x = "Upset (Odds) Occured", title = "Upset Result by Odds of the Favorite") + ylim(1.2,1.8) + facet_wrap(~Round) + my.theme

# Odds of the Underdog
ggplot(data = train, aes(x = odds.upset, y = UMaxOdds)) + geom_boxplot(fill = "orange3") + labs(y = "Odds of Underdog", x = "Upset (Odds) Occured", title = "Upset Result Odds of the Underdog") + ylim(1.85,3.25) + facet_wrap(~Round) + my.theme
```

### Graph Winning Potential
#### First, set up the data...
```{r}
#detach(package:plyr)
library(dplyr)

# Row-proportions: Best.of vs. odds.upset
df.bestof <- as.data.frame(prop.table(table(train$Best.of, train$odds.upset),1))
names(df.bestof) <- c("Best.of", "odds.upset", "upset.freq")
df.bestof <- df.bestof %>% filter(odds.upset == TRUE) %>% select(1,3)

t1<- train %>% group_by(Best.of) %>% filter(odds.upset==FALSE) %>% summarize(chalk.avg = mean(FMaxOdds))
t2<- train %>% group_by(Best.of) %>% filter(odds.upset==TRUE) %>% summarize(dog.avg = mean(UMaxOdds))
df.bestof$chalk.avg.odds <- as.data.frame(t1)[,2]
df.bestof$dog.avg.odds <- as.data.frame(t2)[,2]

df.bestof <- df.bestof %>% mutate(bet.favorite = (1-upset.freq)*chalk.avg.odds, bet.underdog = (upset.freq*dog.avg.odds))


# Row-proportions: Surface vs. odds.upset
df.surface <- as.data.frame(prop.table(table(train$Surface, train$odds.upset),1))
names(df.surface) <- c("Surface", "odds.upset", "upset.freq")
df.surface <- df.surface %>% filter(odds.upset == TRUE) %>% select(1,3)

t1<- train %>% group_by(Surface) %>% filter(odds.upset==FALSE) %>% summarize(chalk.avg = mean(FMaxOdds))
t2<- train %>% group_by(Surface) %>% filter(odds.upset==TRUE) %>% summarize(dog.avg = mean(UMaxOdds))
df.surface$chalk.avg.odds <- as.data.frame(t1)[,2]
df.surface$dog.avg.odds <- as.data.frame(t2)[,2]

df.surface <- df.surface %>% mutate(bet.favorite = (1-upset.freq)*chalk.avg.odds, bet.underdog = (upset.freq*dog.avg.odds))


# Row-proportions: Round vs. odds.upset
df.round <- as.data.frame(prop.table(table(train$Round, train$odds.upset),1))
names(df.round) <- c("Round", "odds.upset", "upset.freq")
df.round <- df.round %>% filter(odds.upset == TRUE) %>% select(1,3)

t1<- train %>% group_by(Round) %>% filter(odds.upset==FALSE) %>% summarize(chalk.avg = mean(FMaxOdds))
t2<- train %>% group_by(Round) %>% filter(odds.upset==TRUE) %>% summarize(dog.avg = mean(UMaxOdds))
df.round$chalk.avg.odds <- as.data.frame(t1)[,2]
df.round$dog.avg.odds <- as.data.frame(t2)[,2]

df.round <- df.round %>% mutate(bet.favorite = (1-upset.freq)*chalk.avg.odds, bet.underdog = (upset.freq*dog.avg.odds))


# Row-proportions: Series vs. odds.upset
df.series <- as.data.frame(prop.table(table(train$Series,train$odds.upset),1))
names(df.series) <- c("Series", "odds.upset", "upset.freq")
df.series <- df.series %>% filter(odds.upset == TRUE) %>% select(1,3)

t1<- train %>% group_by(Series) %>% filter(odds.upset==FALSE) %>% summarize(chalk.avg = mean(FMaxOdds))
t2<- train %>% group_by(Series) %>% filter(odds.upset==TRUE) %>% summarize(dog.avg = mean(UMaxOdds))
df.series$chalk.avg.odds <- as.data.frame(t1)[,2]
df.series$dog.avg.odds <- as.data.frame(t2)[,2]

df.series <- df.series %>% mutate(bet.favorite = (1-upset.freq)*chalk.avg.odds, bet.underdog = (upset.freq*dog.avg.odds))


# Average dog.proft & chalk.proft for horizontal lines
train.1 <- train %>% filter(odds.upset==FALSE)
train.2 <- train %>% filter(odds.upset==TRUE)
overall.chalk.profit <- mean(train.1$FMaxOdds)*mean(1-as.numeric(train$odds.upset))
overall.dog.profit <- mean(train.2$UMaxOdds)*mean(as.numeric(train$odds.upset))

df.bestof.1 <- gather(df.bestof,key="chalk.dog",value="profit",c(5,6))
df.round.1 <- gather(df.round,key="chalk.dog",value="profit",c(5,6))
df.surface.1 <- gather(df.surface,key="chalk.dog",value="profit",c(5,6))
df.series.1 <- gather(df.series,key="chalk.dog",value="profit",c(5,6))

names(df.bestof) <- c("Best.of","upset.freq","fav.avg.odds","dog.avg.odds","fav.potential","dog.potential")
names(df.round) <- c("Round","upset.freq","fav.avg.odds","dog.avg.odds","fav.potential","dog.potential")
names(df.surface) <- c("Surface","upset.freq","fav.avg.odds","dog.avg.odds","fav.potential","dog.potential")
names(df.series) <- c("Series","upset.freq","fav.avg.odds","dog.avg.odds","fav.potential","dog.potential")

df.bestof
df.round
df.surface
df.series
```
#### Payout-Potential Graphs
```{r}
# theme
my.theme <- theme(axis.text.x=element_text(face="bold",color="cyan4", size=8, angle=30), axis.text.y=element_text(face="bold",color="cyan4",size=8), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5), legend.title=element_blank())

# color pallete
my.colors <- as.character(c("chalk.profit"="lightskyblue3", "dog.profit"="orange3"))
my.color.scale <- scale_colour_manual(name="my.colors", values=my.colors)


# Best OF
ggplot(data=df.bestof.1, aes(x=Best.of, y=profit)) + geom_point(aes(color = chalk.dog),size=8) + my.color.scale + geom_hline(yintercept = overall.dog.profit, linetype = "longdash", color = "orange3") + geom_hline(yintercept = overall.chalk.profit, linetype = "dotdash", color = "lightskyblue3") + geom_hline(yintercept=1, color = "grey") + labs(title ="Net Payout Potential by Strategy & Number of Games", y="Payout Potential (per $1 bet)", x = "Possible Games per Match") + theme(axis.text.x=element_text(face="bold",color="cyan4", size=8), axis.text.y=element_text(face="bold",color="cyan4",size=8), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5), legend.title=element_blank()) + geom_text(x = 3.3, y = .925, label = "Overall Underdog", size = 3) + geom_text(x=3.3, y=.995, label = "Overall Favorite", size = 3) + scale_y_continuous(limits = c(.8,1.05), breaks = seq(.8,1.1,.05))


# SURFACE
ggplot(data=df.surface.1, aes(x=Surface, y=profit)) + geom_point(aes(color = chalk.dog),size=8) + my.color.scale + geom_hline(yintercept = overall.dog.profit, linetype = "longdash", color = "orange3") + geom_hline(yintercept = overall.chalk.profit, linetype = "dotdash", color = "lightskyblue3") + geom_hline(yintercept=1, color = "grey") + labs(title ="Net Payout Potential by Strategy & Court Type", y="Payout Potential (per $1 bet)", x = "Surface") + theme(axis.text.x=element_text(face="bold",color="cyan4", size=8), axis.text.y=element_text(face="bold",color="cyan4",size=8), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5), legend.title=element_blank()) + geom_text(x = 3.3, y = .925, label = "Overall Underdog", size = 3) + geom_text(x=3.3, y=.995, label = "Overall Favorite", size = 3) + scale_y_continuous(limits = c(.9,1.02), breaks = seq(.9,1.2,.02))


# ROUND
ggplot(data=df.round.1, aes(x=Round, y=profit)) + geom_point(aes(color = chalk.dog),size=8) + my.color.scale + geom_hline(yintercept = overall.dog.profit, linetype = "longdash", color = "orange3") + geom_hline(yintercept = overall.chalk.profit, linetype = "dotdash", color = "lightskyblue3") + geom_hline(yintercept=1, color = "grey") + labs(title ="Net Payout Potential by Strategy & Round", y="Payout Potential (per $1 bet)", x = "Round") + my.theme + geom_text(x = 4.3, y = .93, label = "Overall Underdog", size = 3) + geom_text(x=4.3, y=1, label = "Overall Favorite", size = 3) + scale_y_continuous(limits = c(.8,1.05), breaks = seq(.8,1.5,.04))


# SERIES
ggplot(data=df.series.1, aes(x=Series, y=profit)) + geom_point(aes(color = chalk.dog),size=8) + my.color.scale + geom_hline(yintercept = overall.dog.profit, linetype = "longdash", color = "orange3") + geom_hline(yintercept = overall.chalk.profit, linetype = "dotdash", color = "lightskyblue3") + geom_hline(yintercept=1, color = "grey") + labs(title ="Net Payout Potential by Strategy & Series", y="Payout Potential (per $1 bet)", x = "Series") + my.theme + geom_text(x = 8.7, y = .93, label = "Overall Underdog", size = 3) + geom_text(x=8.7, y=1, label = "Overall Favorite", size = 3) + scale_y_continuous(limits = c(.7,1.10), breaks = seq(.7,1.10,.05))
```



### Models
```{r}
model.1 <- glm(odds.upset~rank.dif + odds.dif + Best.of + Round + Surface, family = binomial, data = train)
summary(model.1)
```



# Gambling Strategy Simulation

### Tree Strategy
```{r}
nick.strategy <- function(df, wager = 10){

  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # Vector of decision requirements - series
  bet.fav.series <- c("Grand Slam", "International Series", "Masters Cup")
  bet.dog.series <- c("Masters 1000")
  # Vector of baseline betting requirements - series
  bet.series <- c(bet.fav.series, bet.dog.series, "ATP250", "ATP500", "Masters")
  
  # Vector of decision requirements - game
  bet.fav.games <- c(5)
  bet.dog.games <- c()
  # Vector of baseline betting requirements - game
  bet.games <- c(bet.fav.games, 3)
  
  # Vector of decision requirements - surface
  bet.fav.surface <- c("Grass")
  bet.dog.surface <- c("Carpet")
  # Vector of baseline betting requirements - surface
  bet.surface <- c(bet.fav.surface, bet.dog.surface, "Hard", "Clay")
  
  # Vector of decision requirements - round
  bet.fav.round <- c("4th Round", "The Final")
  bet.dog.round <- c()
  # Vector of baseline betting requirements - round
  bet.round <- c(bet.fav.round,"Quarterfinals", "Semifinals")
  
  
  # Loop for betting decisions
  for(i in 1:nrow(df)){
    
    # Match must meet all baseline betting requirements
    if((df$Series[i] %in% bet.series) & (df$Best.of[i] %in% bet.games) & (df$Surface[i] %in% bet.surface) & (df$Round[i] %in% bet.round)){
    
      
      # BET FAVORITE if any favorite decision requirement met
      if((df$Series[i] %in% bet.fav.series) | (df$Best.of[i] %in% bet.fav.games) | (df$Surface[i] %in% bet.fav.surface) | (df$Round[i] %in% bet.fav.round)){
        
        # place wager on FMaxOdds
        
        # if win
        if(df$odds.upset[i] == FALSE){
          bet.result <- wager * (df$FMaxOdds[i]-1)
          
          net <- net + bet.result
          wins <- wins + 1
          tries <- tries + 1
        }
        
        #if loss
        if(df$odds.upset[i] == TRUE){
          bet.result <- -1*wager
          
          net <- net + bet.result
          tries <- tries + 1
        }
      }
      
      # BET UNDERDOG if any underdog decision requirement met
      if((df$Series[i] %in% bet.dog.series) | (df$Best.of[i] %in% bet.dog.games) | (df$Surface[i] %in% bet.dog.surface) | (df$Round[i] %in% bet.dog.round)){
        
        # place wager on UMaxOdds
        
        # if win
        if(df$odds.upset[i] == TRUE){
          bet.result <- wager * (df$UMaxOdds[i]-1)
          
          net <- net + bet.result
          wins <- wins + 1
          tries <- tries + 1
        }
        
        #if loss
        if(df$odds.upset[i] == FALSE){
          bet.result <- -1*wager
          
          net <- net + bet.result
          tries <- tries + 1
        }
      }
    }
  }
  
  
  # RETURN LIST
  out <- list(net, wins, tries)
  names(out) <- c("net","wins","tries")
  return(out)
}
```

### Loop simulation through 2015, 2016, 2017
```{r}
YEARS <- unique(test$year)
Net <- rep(NA, times=3)
Wins <- rep(NA, times=3)
Tries <- rep(NA, times=3)

sim <- data.frame(YEARS, Net, Wins, Tries)

for(i in 1:length(YEARS)){
  season <- test %>% filter(year == YEARS[i])
  s.result <- nick.strategy(season, wager = 10)
  
  sim$Net[i] <- s.result[[1]]
  sim$Wins[i] <- s.result[[2]]
  sim$Tries[i] <- s.result[[3]]
}
names(sim)[1] <- "Year"
sim
```

## Graph of Simulation Results
```{r}
# theme
my.theme <- theme(axis.text.x=element_text(face="bold",color="cyan4", size=10), axis.text.y=element_text(face="bold",color="cyan4",size=10), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5), legend.title=element_blank(), plot.subtitle = element_text(face="bold", color="royalblue4", size=10, hjust=.5))

# ggplot
ggplot(data=sim, aes(x=Year, y=Net)) + stat_summary(fun.y=mean, geom="bar", fill="orange3") + labs(x="Year", y="Net Result ($10 Wagers)", title="Conditional Betting Strategy", subtitle="From Most Favorable Book") + my.theme
```
