---
title: "Baseline Gambling Simulations"
author: "Nick Kurtansky"
date: "11/25/2017"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of this is to provide the results to the two baseline strategies:
  1. Always betting the favorite (average book)
  2. Always betting the underdog (average book)
  3. Randomly betting the underdog or favorite (average book)
  4. Always betting the higher ranked player (average book)
  5. Always betting the lower ranked player (average book)
  6. Always betting the favorite (highest book)
  7. Always betting the underdog (highest book)
  8. Randomly betting the underdog or favorite (highest book)
  9. Always betting the higher ranked player (highest book)
  10. Always betting the lower ranked player (highest book)
  
I will run this simulation seperately for each of the years from 2001 through 2017 in order to look for trends that may provide insight in how the oddsmakers have adapted over the years.

```{r include = FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(knitr)
```

### load in data
```{r}
load("data/favorite.data.RData")
data <- favorite.data

#head(data,100)
#str(data)
```

### Clean Data
```{r}
# only include matches that sports books took bets on
data <- data %>% filter((FcountOdds > 0) & (UcountOdds > 0))

# only include matches between players who have a rank
data <- data %>% filter((!is.na(URank))&(!is.na(FRank)))

# new column for absolute rank difference variable: rank.dif
data <- data %>% mutate(rank.dif = URank - FRank)

# new column for absolute odds difference variable: odds.dif
data <- data %>% mutate(abs.odds.dif = avgU - avgF)

str(data)
#data %>% filter(year == 2008)
```

### What is the raw probability of an upset?
```{r}
sum(data$odds.upset/nrow(data))
```

Upset occurs with probability 0.2999496... Hence, odds have a success rate p = 0.7000504


# Simulations

#### Outcome Matrix
Includes:
  Year, strategy, net profit, wins, tries
```{r}
Year <- c(2001:2017)
B1 <- rep("bet fav - avg odds", times=17)
B2 <- rep("bet dog - avg odds", times=17)
B3 <- rep("bet random - avg odds", times=17)
B4 <- rep("bet higher rank - avg odds", times=17)
B5 <- rep("bet lower rank - avg odds",times=17)
B6 <- rep("bet fav - best odds", times=17)
B7 <- rep("bet dog - best odds", times=17)
B8 <- rep("bet random - best odds", times=17)
B9 <- rep("bet higher rank - best odds", times=17)
B10 <- rep("bet lower rank - best odds",times=17)
Net <- rep(NA, times=17)
Wins <- rep(NA, times=17)
Tries <- rep(NA, times=17)

df1 <- data.frame(Year, B1, Net, Wins, Tries)
df2 <- data.frame(Year, B2, Net, Wins, Tries)
df3 <- data.frame(Year, B3, Net, Wins, Tries)
df4 <- data.frame(Year, B4, Net, Wins, Tries)
df5 <- data.frame(Year, B5, Net, Wins, Tries)
df6 <- data.frame(Year, B6, Net, Wins, Tries)
df7 <- data.frame(Year, B7, Net, Wins, Tries)
df8 <- data.frame(Year, B8, Net, Wins, Tries)
df9 <- data.frame(Year, B9, Net, Wins, Tries)
df10 <- data.frame(Year, B10, Net, Wins, Tries)
```


## Baseline Strategy 1: always bet favorite average book

#### Function for single-year simulation
```{r}
bet.favorite <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){
    
    # if chalk (win)
    if(df$odds.upset[i] == FALSE){
      bet.result <- wager * (df$avgF[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if upset (loss)
    if(df$odds.upset[i] == TRUE){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.favorite(season, wager = 10)
  
  df1$Net[i] <- s.result[[1]]
  df1$Wins[i] <- s.result[[2]]
  df1$Tries[i] <- s.result[[3]]
}
names(df1)[2] <- "Strategy"
df1
```


NOTE OF CONCERN:
The year 2013 looks wrong. I know Will noticed some red-flags last week and made some adjustments, but I am still seeing issues here.

## Baseline Strategy 2: always bet underdog (average book)

#### Function for single-year simulation
```{r}
bet.underdog <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){
    
    # if upset (win)
    if(df$odds.upset[i] == TRUE){
      bet.result <- wager * (df$avgU[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if chalk (loss)
    if(df$odds.upset[i] == FALSE){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.underdog(season, wager = 10)
  
  df2$Net[i] <- s.result[[1]]
  df2$Wins[i] <- s.result[[2]]
  df2$Tries[i] <- s.result[[3]]
}
names(df2)[2] <- "Strategy"
df2
```

NOTE OF CONCERN: More issues with 2013; There is no chance we should be seeing $230,000 payout on $10 underdog bets. Also, I wonder about 2012. It is the only year we notice a profit... Hmm...


## Baseline Strategy 3: random choice of underdog/favorite (average book)

#### Function for single-year simulation
```{r}
bet.random <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){
    
    choice <- sample(c("fav","dog"),size=1)
    
    if(choice == "dog"){

      # if upset (win)
      if(df$odds.upset[i] == TRUE){
        bet.result <- wager * (df$avgU[i]-1)
        
        net <- net + bet.result

        wins <- wins + 1
        tries <- tries + 1
      }
      
      # if chalk (loss)
      if(df$odds.upset[i] == FALSE){
        bet.result <- -1*wager
        
        net <- net + bet.result
        tries <- tries + 1
      }
    }
    
    if(choice == "fav"){

      # if chalk (win)
      if(df$odds.upset[i] == FALSE){
        bet.result <- wager * (df$avgF[i]-1)
  
        net <- net + bet.result
        wins <- wins + 1
        tries <- tries + 1
      }
      
      # if upset (loss)
      if(df$odds.upset[i] == TRUE){
        bet.result <- -1*wager
        
        net <- net + bet.result
        tries <- tries + 1
      }
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.random(season, wager = 10)
  
  df3$Net[i] <- s.result[[1]]
  df3$Wins[i] <- s.result[[2]]
  df3$Tries[i] <- s.result[[3]]
}
names(df3)[2] <- "Strategy"
df3
```
More successful than betting all underdogs. Less successful than betting all favorites. We still lose money in the end.


## Baseline Strategy 4: always bet the higher ranked player (average book)

#### Function for single-year simulation
```{r}
bet.highrank <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){

    if(df$rank.upset[i] == FALSE){

      # if 'chalk' and higher rank (win)
      if(df$odds.upset[i]==FALSE){
        bet.result <- wager * (df$avgF[i]-1)
        
        net <- net + bet.result
        wins <- wins + 1
        tries <- tries + 1
      }
    
      # if 'dog' and higher rank (win)
      if(df$odds.upset[i]==TRUE){
        bet.result <- wager * (df$avgU[i]-1)
        
        net <- net + bet.result
        wins <- wins + 1
        tries <- tries + 1
      }
    }
    
    if(df$rank.upset[i] == TRUE){

      # if 'chalk' and lower rank (loss)
      if(df$odds.upset[i] == FALSE){
        bet.result <- -1*wager
        
        net <- net + bet.result
        tries <- tries + 1
      }
    
      # if 'dog' and lower rank (loss)
      if(df$odds.upset[i] == TRUE){
        bet.result <- -1*wager
        
        net <- net + bet.result
        tries <- tries + 1
      }
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.highrank(season, wager = 10)
  
  df4$Net[i] <- s.result[[1]]
  df4$Wins[i] <- s.result[[2]]
  df4$Tries[i] <- s.result[[3]]
}
names(df4)[2] <- "Strategy"
df4
```
I'm concerned with the amount of profit we make off this strategy. It seems far to unlikely that it would be this easy. I suspect something is wrong with the variable 'upset' in the original RData, which I have converted here to 'rank.upset'.


## Baseline Strategy 5: always bet the lower ranked player (average book)

#### Function for single-year simulation
```{r}
bet.lowrank <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){
    
    # if 'chalk' and lower rank (win)
    if((df$rank.upset[i] == TRUE)&(df$odds.upset[i]==FALSE)){
      bet.result <- wager * (df$avgF[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if 'dog' and lower rank (win)
    if((df$rank.upset[i] == TRUE)&(df$odds.upset[i]==TRUE)){
      bet.result <- wager * (df$avgU[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if 'chalk' and higher rank (loss)
    if((df$rank.upset[i] == FALSE)&(df$odds.upset[i] == FALSE)){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
    
    # if 'dog' and higher rank (win)
    if((df$rank.upset[i] == FALSE)&(df$odds.upset[i] == TRUE)){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.lowrank(season, wager = 10)
  
  df5$Net[i] <- s.result[[1]]
  df5$Wins[i] <- s.result[[2]]
  df5$Tries[i] <- s.result[[3]]
}
names(df5)[2] <- "Strategy"
df5
```
It makes sence that you wouldn't be successsful always putting your money on the lower ranked player.


## Baseline Strategy 6: always bet favorite (highest book)

#### Function for single-year simulation
```{r}
bet.favorite.h <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){
    
    # if chalk (win)
    if(df$odds.upset[i] == FALSE){
      bet.result <- wager * (df$FMaxOdds[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if upset (loss)
    if(df$odds.upset[i] == TRUE){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.favorite.h(season, wager = 10)
  
  df6$Net[i] <- s.result[[1]]
  df6$Wins[i] <- s.result[[2]]
  df6$Tries[i] <- s.result[[3]]
}
names(df6)[2] <- "Strategy"
df6
```



## Baseline Strategy 7: always bet underdog (highest book)

#### Function for single-year simulation
```{r}
bet.underdog.h <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){
    
    # if upset (win)
    if(df$odds.upset[i] == TRUE){
      bet.result <- wager * (df$UMaxOdds[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if chalk (loss)
    if(df$odds.upset[i] == FALSE){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.underdog.h(season, wager = 10)
  
  df7$Net[i] <- s.result[[1]]
  df7$Wins[i] <- s.result[[2]]
  df7$Tries[i] <- s.result[[3]]
}
names(df7)[2] <- "Strategy"
df7
```



## Baseline Strategy 8: random choice of underdog/favorite (highest book)

#### Function for single-year simulation
```{r}
bet.random.h <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){
    
    choice <- sample(c("fav","dog"),size=1)
    
    if(choice == "dog"){

      # if upset (win)
      if(df$odds.upset[i] == TRUE){
        bet.result <- wager * (df$UMaxOdds[i]-1)
        
        net <- net + bet.result

        wins <- wins + 1
        tries <- tries + 1
      }
      
      # if chalk (loss)
      if(df$odds.upset[i] == FALSE){
        bet.result <- -1*wager
        
        net <- net + bet.result
        tries <- tries + 1
      }
    }
    
    if(choice == "fav"){

      # if chalk (win)
      if(df$odds.upset[i] == FALSE){
        bet.result <- wager * (df$FMaxOdds[i]-1)
  
        net <- net + bet.result
        wins <- wins + 1
        tries <- tries + 1
      }
      
      # if upset (loss)
      if(df$odds.upset[i] == TRUE){
        bet.result <- -1*wager
        
        net <- net + bet.result
        tries <- tries + 1
      }
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.random.h(season, wager = 10)
  
  df8$Net[i] <- s.result[[1]]
  df8$Wins[i] <- s.result[[2]]
  df8$Tries[i] <- s.result[[3]]
}
names(df8)[2] <- "Strategy"
df8
```




## Baseline Strategy 9: always bet the higher ranked player (highest book)

#### Function for single-year simulation
```{r}
bet.highrank.h <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){
    
    # if 'chalk' and higher rank (win)
    if((df$rank.upset[i] == FALSE)&(df$odds.upset[i]==FALSE)){
      bet.result <- wager * (df$FMaxOdds[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if 'dog' and higher rank (win)
    if((df$rank.upset[i] == FALSE)&(df$odds.upset[i]==TRUE)){
      bet.result <- wager * (df$UMaxOdds[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if 'chalk' and lower rank (loss)
    if((df$rank.upset[i] == TRUE)&(df$odds.upset[i] == FALSE)){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
    
    # if 'dog' and lower rank (loss)
    if((df$rank.upset[i] == TRUE)&(df$odds.upset[i] == TRUE)){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.highrank.h(season, wager = 10)
  
  df9$Net[i] <- s.result[[1]]
  df9$Wins[i] <- s.result[[2]]
  df9$Tries[i] <- s.result[[3]]
}
names(df9)[2] <- "Strategy"
df9
```


## Baseline Strategy 10: always bet the lower ranked player (highest book)

#### Function for single-year simulation
```{r}
bet.lowrank.h <- function(df, wager = 10){
  # initialize
  net <- 0
  tries <- 0
  wins <- 0
  
  # for each match:
  for(i in 1:nrow(df)){
    
    # if 'chalk' and lower rank (win)
    if((df$rank.upset[i] == TRUE)&(df$odds.upset[i]==FALSE)){
      bet.result <- wager * (df$FMaxOdds[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if 'dog' and lower rank (win)
    if((df$rank.upset[i] == TRUE)&(df$odds.upset[i]==TRUE)){
      bet.result <- wager * (df$UMaxOdds[i]-1)
      
      net <- net + bet.result
      wins <- wins + 1
      tries <- tries + 1
    }
    
    # if 'chalk' and higher rank (loss)
    if((df$rank.upset[i] == FALSE)&(df$odds.upset[i] == FALSE)){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
    
    # if 'dog' and higher rank (win)
    if((df$rank.upset[i] == FALSE)&(df$odds.upset[i] == TRUE)){
      bet.result <- -1*wager
      
      net <- net + bet.result
      tries <- tries + 1
    }
  }
  
  # output a list
  out <- list(net, wins, tries)
  names(out) <- c("net","wins","tries")
  return(out)
}
```

#### Loop through all the seasons
```{r}
for(i in 1:length(Year)){
  season <- data %>% filter(year == Year[i])
  s.result <- bet.lowrank.h(season, wager = 10)
  
  df10$Net[i] <- s.result[[1]]
  df10$Wins[i] <- s.result[[2]]
  df10$Tries[i] <- s.result[[3]]
}
names(df10)[2] <- "Strategy"
df10
```

###Create Table for Strategies 6-10

```{r}
#Sum Net Profit, number of wins, number of bets from strategies 6-10
df6.sums <- colSums(df6[,-1:-2])

df7.sums <- colSums(df7[,-1:-2])

df8.sums <- colSums(df8[,-1:-2])

df9.sums <- colSums(df9[,-1:-2])

df10.sums <- colSums(df10[,-1:-2])

```
```{r}
#create column strategy
strategy <- c(6,7,8,9,10)
#create net profit column from net profit columns
net <- c(df6.sums[1], df7.sums[1], df8.sums[1], df9.sums[1], df10.sums[1])
#create number of wins column 
wins <- c(df6.sums[2], df7.sums[2], df8.sums[2], df9.sums[2], df10.sums[2])
#create number of bets column
bets <- c(df6.sums[3], df7.sums[3], df8.sums[3], df9.sums[3], df10.sums[3])
#create win percentage column by dividing number of wins over number of bets
win.percentage <- c(df6.sums[2]/df6.sums[3], df7.sums[2]/df7.sums[3], df8.sums[2]/df8.sums[3], df9.sums[2]/df9.sums[3], df10.sums[2]/df10.sums[3])
#build dataframe from strategy, net, wins, bets and win percentage
df.6.10 <- data.frame(strategy, net, wins, bets, win.percentage)
#creates column names
colnames(df.6.10) <- c("Strategy", "Net Profit", "Number of Wins", "Number of Bets", "Win Percentage")
df.6.10
#builds the table
kable(df.6.10)
```


## View Results
```{r}
Result <- rbind(df1, df2, df3, df4, df5, df6, df7, df8, df9, df10)
names(Result) <- c("Year","Strategy","Net Profit (w/ $10 bets)","# Successes","# Tries")

#Result
boxplot(df1$Net)
boxplot(df2$Net)
boxplot(df3$Net)
boxplot(df4$Net)
boxplot(df5$Net)
boxplot(df6$Net)
boxplot(df7$Net)
boxplot(df8$Net)
boxplot(df9$Net)
boxplot(df10$Net)
```

#### Comments on Results

I am concerned with the data we have for 2013. There surely is something wrong there with the filtered.men.data frame I used from the github. Another note of concern is the year 2012, where betting the underdog every time resulted in net profit. This leads to the ultimate conclusion from the baseline strategies: YOU CANNOT WIN WITH THEM... You will lose in the long run. However, betting the favorite is not quite as devastating as betting the underdog.