---
title: "Simulation using Player Logistic Regressions"
author: "Jacob Schmitter"
date: "November 27, 2017"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment = NA)
library(tidyverse)
library(ggplot2)
library(arm)
library(lubridate)
```

```{r}
load("aggregated_mens_logistic.RData")
```

```{r, warning=FALSE}
#load mens data and format dates column
men.data <- t2
#format date column
dates <- mdy(men.data$Date)
men.data$Date <- dates
#convert variables to numeric and factors
men.data$LRank <- as.numeric(men.data$LRank)
men.data$WRank <- as.numeric(men.data$WRank)
men.data$LPts <- as.numeric(men.data$LPts)
men.data$WPts <- as.numeric(men.data$WPts)
men.data$Surface <- as.factor(men.data$Surface)
men.data$Round <- as.factor(men.data$Round)
men.data$Series <- as.factor(men.data$Series)
men.data$Best.of <- as.factor(men.data$Best.of)

  #remove rows with NaN in average betting odds columns, rank columns, and point columns
  men.data <- men.data[!is.na(men.data$avgW),]
  men.data <- men.data[!is.na(men.data$avgL),]
  men.data <- men.data[!is.na(men.data$avgW),]
  men.data <- men.data[!is.na(men.data$avgL),]
  men.data <- men.data[!is.na(men.data$WRank),]
  men.data <- men.data[!is.na(men.data$LRank),]
  men.data <- men.data[!is.na(men.data$WRank),]
  men.data <- men.data[!is.na(men.data$LRank),]
  men.data <- men.data[!is.na(men.data$WPts),]
  men.data <- men.data[!is.na(men.data$LPts),]
  men.data <- men.data[!is.na(men.data$WPts),]
  men.data <- men.data[!is.na(men.data$LPts),]
```

A potential betting strategy that could work effectively is to be selective in the matches that you bet on, and only bet when you believe that the player has a very high probability of winning the match. A way to do this is use a logistic regression model for a given player and only bet if the model predicts that they have at least a certain probability of winning given different variables such as surface type, rank difference, and odds. 

Below I create a function that simulates betting on the 2015, 2016, and 2017 seasons for a given player and runs a logistic regression model that predicts the probability of the given player of winning the match. The logistic regression model uses the 2001 to 2014 seasons as training data for each player. We will use the logistic regression model to look at different net payoffs if the minimum threshold for the winning probability in our betting strategy is changed.

```{r}
#function that simulates betting on 2015,2016, and 2017 seasons for a given player and runs a logistic regression to predict probability the player wins the match

player.logistic.simulation <- function(playerName, wager=10){
  #filter data frame to games containing a specific player and seperate into training & testing data
  player.train <- filter(men.data, (Winner %in% playerName | Loser %in% playerName) & Date <= "2014-12-31")
  player.test <- filter(men.data, (Winner %in% playerName | Loser %in% playerName) & Date >="2014-12-31")

  #create variable with outcome of game for player
  player.train$Outcome <- (player.train$Winner == playerName)
  player.train$Outcome <- as.numeric(player.train$Outcome)
  player.test$Outcome <- (player.test$Winner == playerName)
  player.test$Outcome <- as.numeric(player.test$Outcome)
  
  #create variable for odds, rank difference, and point difference of player in both testing and training data frames
  for(i in 1:dim(player.train)[1]){
    if(((player.train$Outcome[i] == 1))){
    player.train$Odds[i] <- player.train$avgW[i]
    player.train$RankDif[i] <- player.train$WRank[i]-player.train$LRank[i]
    player.train$PtDif[i] <- player.train$WPts[i] -player.train$LPts[i]
  }
    else{
    player.train$Odds[i] <- player.train$avgL[i]
    player.train$RankDif[i] <- player.train$LRank[i]-player.train$WRank[i]
    player.train$PtDif[i] <- -1*player.train$WPts[i] +player.train$LPts[i]
  }
}
  for(i in 1:dim(player.test)[1]){
    if(((player.test$Outcome[i] == 1))){
      player.test$Odds[i] <- player.test$avgW[i]
      player.test$RankDif[i] <- player.test$WRank[i]-player.test$LRank[i]
      player.test$PtDif[i] <- player.test$WPts[i] -player.test$LPts[i]
    }
    else{
      player.test$Odds[i] <- player.test$avgL[i]
      player.test$RankDif[i] <- player.test$LRank[i]-player.test$WRank[i]
      player.test$PtDif[i] <- -1*player.test$WPts[i] +player.test$LPts[i]
    }
  }
  
  

  #create logistic regression model using training data
  player.glm <- bayesglm(Outcome ~ Surface+Odds+RankDif+Best.of+Series+PtDif, family=binomial, data=player.train,drop.unused.levels   = FALSE)
  
  #format testing data to be able to input into predict.glm
  player.test <- subset(player.test, select=c("year","Surface","Odds","Outcome","Series","RankDif","PtDif","Best.of"))
  prob.win <- predict.glm(player.glm, newdata=subset(player.test,select=c("Surface","Odds","Series","RankDif","PtDif","Best.of")),type = "response")
  

  #attach predicted probability of winning
  player.test$prob <- as.numeric(prob.win)
  player.test$year <- as.factor(player.test$year)
  
  #payoff of match if we bet
  player.test <- mutate(player.test, payoff= Outcome*wager*(Odds-1))
  player.test$payoff[player.test$Outcome==0] <- -1*wager


  #output information regarding betting simulation
  out <- player.test
  return(out)
  }
```

Since the model is based solely on historic performance, it will only work effectively on players who have a larger history for the model to be based on. Therefore, I will run the simulation on all players in the current top 15 who are ages 29 and above. Below I run the simulation for these players and display the results of the simulations for if I had only bet on games in which the player had at least a 0.85, 0.875, or 0.9 probability of winning from the logistic regression's predictions.

```{r,warning=FALSE, echo = FALSE}
#betting on all individuals in the current top 15 that are 29 years of age or older for the years 2015,2016,2017

#running simulation for each player and organizing data to analyze for 0.85,0.875, and 0.9 probability minimums
#nadal
nadal <- player.logistic.simulation("Nadal R.", wager = 10)
# 0.85 odds
nadal <- filter(nadal, prob >= 0.85)
nadal.net.85 <- sum(nadal$payoff)
nadal.wins.85 <- sum(nadal$Outcome)
nadal.tries.85 <- nrow(nadal)
nadal.year.net.85 <- nadal %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.875 odds
nadal <- filter(nadal, prob >= 0.875)
nadal.net.875 <- sum(nadal$payoff)
nadal.wins.875 <- sum(nadal$Outcome)
nadal.tries.875 <- nrow(nadal)
nadal.year.net.875 <- nadal %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.9 odds
nadal <- filter(nadal, prob >= 0.9)
nadal.net.9 <- sum(nadal$payoff)
nadal.wins.9 <- sum(nadal$Outcome)
nadal.tries.9 <- nrow(nadal)
nadal.year.net.9 <- nadal %>% group_by(year) %>% summarize(net = sum(payoff))

#federer
federer <- player.logistic.simulation("Federer R.", wager = 10)
# 0.85 odds
federer <- filter(federer, prob >= 0.85)
federer.net.85 <- sum(federer$payoff)
federer.wins.85 <- sum(federer$Outcome)
federer.tries.85 <- nrow(federer)
federer.year.net.85 <- federer %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.875 odds
federer <- filter(federer, prob >= 0.875)
federer.net.875 <- sum(federer$payoff)
federer.wins.875 <- sum(federer$Outcome)
federer.tries.875 <- nrow(federer)
federer.year.net.875 <- federer %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.9 odds
federer <- filter(federer, prob >= 0.9)
federer.net.9 <- sum(federer$payoff)
federer.wins.9 <- sum(federer$Outcome)
federer.tries.9 <- nrow(federer)
federer.year.net.9 <- federer %>% group_by(year) %>% summarize(net = sum(payoff))

#djokovic
djokovic <- player.logistic.simulation("Djokovic N.", wager = 10)
# 0.85 odds
djokovic <- filter(djokovic, prob >= 0.85)
djokovic.net.85 <- sum(djokovic$payoff)
djokovic.wins.85 <- sum(djokovic$Outcome)
djokovic.tries.85 <- nrow(djokovic)
djokovic.year.net.85 <- djokovic %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.875 odds
djokovic <- filter(djokovic, prob >= 0.875)
djokovic.net.875 <- sum(djokovic$payoff)
djokovic.wins.875 <- sum(djokovic$Outcome)
djokovic.tries.875 <- nrow(djokovic)
djokovic.year.net.875 <- djokovic %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.9 odds
djokovic <- filter(djokovic, prob >= 0.9)
djokovic.net.9 <- sum(djokovic$payoff)
djokovic.wins.9 <- sum(djokovic$Outcome)
djokovic.tries.9 <- nrow(djokovic)
djokovic.year.net.9 <- djokovic %>% group_by(year) %>% summarize(net = sum(payoff))

#wawrinka
wawrinka <- player.logistic.simulation("Wawrinka S.", wager = 10)
# 0.85 odds
wawrinka <- filter(wawrinka, prob >= 0.85)
wawrinka.net.85 <- sum(wawrinka$payoff)
wawrinka.wins.85 <- sum(wawrinka$Outcome)
wawrinka.tries.85 <- nrow(wawrinka)
wawrinka.year.net.85 <- wawrinka %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.875 odds
wawrinka <- filter(wawrinka, prob >= 0.875)
wawrinka.net.875 <- sum(wawrinka$payoff)
wawrinka.wins.875 <- sum(wawrinka$Outcome)
wawrinka.tries.875 <- nrow(wawrinka)
wawrinka.year.net.875 <- wawrinka %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.9 odds
wawrinka <- filter(wawrinka, prob >= 0.9)
wawrinka.net.9 <- sum(wawrinka$payoff)
wawrinka.wins.9 <- sum(wawrinka$Outcome)
wawrinka.tries.9 <- nrow(wawrinka)
wawrinka.year.net.9 <- wawrinka %>% group_by(year) %>% summarize(net = sum(payoff))

#delpotro
delpotro <- player.logistic.simulation("Del Potro J.M.", wager = 10)
# 0.85 odds
delpotro <- filter(delpotro, prob >= 0.85)
delpotro.net.85 <- sum(delpotro$payoff)
delpotro.wins.85 <- sum(delpotro$Outcome)
delpotro.tries.85 <- nrow(delpotro)
delpotro.year.net.85 <- delpotro %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.875 odds
delpotro <- filter(delpotro, prob >= 0.875)
delpotro.net.875 <- sum(delpotro$payoff)
delpotro.wins.875 <- sum(delpotro$Outcome)
delpotro.tries.875 <- nrow(delpotro)
delpotro.year.net.875 <- delpotro %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.9 odds
delpotro <- filter(delpotro, prob >= 0.9)
delpotro.net.9 <- sum(delpotro$payoff)
delpotro.wins.9 <- sum(delpotro$Outcome)
delpotro.tries.9 <- nrow(delpotro)
delpotro.year.net.9 <- delpotro %>% group_by(year) %>% summarize(net = sum(payoff))

#cilic
cilic <- player.logistic.simulation("Cilic M.", wager = 10)
# 0.85 odds
cilic <- filter(cilic, prob >= 0.85)
cilic.net.85 <- sum(cilic$payoff)
cilic.wins.85 <- sum(cilic$Outcome)
cilic.tries.85 <- nrow(cilic)
cilic.year.net.85 <- cilic %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.875 odds
cilic <- filter(cilic, prob >= 0.875)
cilic.net.875 <- sum(cilic$payoff)
cilic.wins.875 <- sum(cilic$Outcome)
cilic.tries.875 <- nrow(cilic)
cilic.year.net.875 <- cilic %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.9 odds
cilic <- filter(cilic, prob >= 0.9)
cilic.net.9 <- sum(cilic$payoff)
cilic.wins.9 <- sum(cilic$Outcome)
cilic.tries.9 <- nrow(cilic)
cilic.year.net.9 <- cilic %>% group_by(year) %>% summarize(net = sum(payoff))

#queery
queery <- player.logistic.simulation("Querrey S.", wager = 10)
# 0.85 odds
queery <- filter(queery, prob >= 0.85)
queery.net.85 <- sum(queery$payoff)
queery.wins.85 <- sum(queery$Outcome)
queery.tries.85 <- nrow(queery)
queery.year.net.85 <- queery %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.875 odds
queery <- filter(queery, prob >= 0.875)
queery.net.875 <- sum(queery$payoff)
queery.wins.875 <- sum(queery$Outcome)
queery.tries.875 <- nrow(queery)
queery.year.net.875 <- queery %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.9 odds
queery <- filter(queery, prob >= 0.9)
queery.net.9 <- sum(queery$payoff)
queery.wins.9 <- sum(queery$Outcome)
queery.tries.9 <- nrow(queery)
queery.year.net.9 <- queery %>% group_by(year) %>% summarize(net = sum(payoff))

#anderson
anderson <- player.logistic.simulation("Anderson K.", wager = 10)
# 0.85 odds
anderson <- filter(anderson, prob >= 0.85)
anderson.net.85 <- sum(anderson$payoff)
anderson.wins.85 <- sum(anderson$Outcome)
anderson.tries.85 <- nrow(anderson)
anderson.year.net.85 <- anderson %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.875 odds
anderson <- filter(anderson, prob >= 0.875)
anderson.net.875 <- sum(anderson$payoff)
anderson.wins.875 <- sum(anderson$Outcome)
anderson.tries.875 <- nrow(anderson)
anderson.year.net.875 <- anderson %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.9 odds
anderson <- filter(anderson, prob >= 0.9)
anderson.net.9 <- sum(anderson$payoff)
anderson.wins.9 <- sum(anderson$Outcome)
anderson.tries.9 <- nrow(anderson)
anderson.year.net.9 <- anderson %>% group_by(year) %>% summarize(net = sum(payoff))

#tsonga
tsonga <- player.logistic.simulation("Tsonga J.W.", wager = 10)
# 0.85 odds
tsonga <- filter(tsonga, prob >= 0.85)
tsonga.net.85 <- sum(tsonga$payoff)
tsonga.wins.85 <- sum(tsonga$Outcome)
tsonga.tries.85 <- nrow(tsonga)
tsonga.year.net.85 <- tsonga %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.875 odds
tsonga <- filter(tsonga, prob >= 0.875)
tsonga.net.875 <- sum(tsonga$payoff)
tsonga.wins.875 <- sum(tsonga$Outcome)
tsonga.tries.875 <- nrow(tsonga)
tsonga.year.net.875 <- tsonga %>% group_by(year) %>% summarize(net = sum(payoff))
# 0.9 odds
tsonga <- filter(tsonga, prob >= 0.9)
tsonga.net.9 <- sum(tsonga$payoff)
tsonga.wins.9 <- sum(tsonga$Outcome)
tsonga.tries.9 <- nrow(tsonga)
tsonga.year.net.9 <- tsonga %>% group_by(year) %>% summarize(net = sum(payoff))
```

```{r, echo = FALSE}
#create data frame to display all betting simulations together
#0.85 odds
net.85 <- c(nadal.net.85,federer.net.85,djokovic.net.85,wawrinka.net.85,delpotro.net.85,cilic.net.85,queery.net.85,anderson.net.85,tsonga.net.85)
wins.85 <- c(nadal.wins.85,federer.wins.85,djokovic.wins.85,wawrinka.wins.85,delpotro.wins.85,cilic.wins.85,queery.wins.85,anderson.wins.85,tsonga.wins.85)
tries.85 <- c(nadal.tries.85,federer.tries.85,djokovic.tries.85,wawrinka.tries.85,delpotro.tries.85,cilic.tries.85,queery.tries.85,anderson.tries.85,tsonga.tries.85)

df.85 <- data.frame("Net"= net.85,"Wins"=wins.85,"Tries"=tries.85)
df.85["Total" ,] <- colSums(df.85)
df.85 <- mutate(df.85, "Win Percentage" = Wins/Tries)
row.names(df.85) <- c("Rafael Nadal","Roger Federer","Novak Djokovic","Stan Wawrinka","Juan Martin del Potro","Marin Cilic","Sam Querrey","Kevin Anderson","Jo-Wilfried Tsonga","Total")

#Look at yearly winnings
year.net.85 <- rbind(nadal.year.net.85,federer.year.net.85,djokovic.year.net.85,wawrinka.year.net.85,delpotro.year.net.85,cilic.year.net.85,queery.year.net.85,anderson.year.net.85,tsonga.year.net.85)
year.net.85 <- year.net.85 %>% group_by(year) %>% summarise(net=sum(net))

#0.875 odds
net.875 <- c(nadal.net.875,federer.net.875,djokovic.net.875,wawrinka.net.875,delpotro.net.875,cilic.net.875,queery.net.875,anderson.net.875,tsonga.net.875)
wins.875 <- c(nadal.wins.875,federer.wins.875,djokovic.wins.875,wawrinka.wins.875,delpotro.wins.875,cilic.wins.875,queery.wins.875,anderson.wins.875,tsonga.wins.875)
tries.875 <- c(nadal.tries.875,federer.tries.875,djokovic.tries.875,wawrinka.tries.875,delpotro.tries.875,cilic.tries.875,queery.tries.875,anderson.tries.875,tsonga.tries.875)

df.875 <- data.frame("Net"= net.875,"Wins"=wins.875,"Tries"=tries.875)
df.875["Total" ,] <- colSums(df.875)
df.875 <- mutate(df.875, "Win Percentage" = Wins/Tries)
row.names(df.875) <- c("Rafael Nadal","Roger Federer","Novak Djokovic","Stan Wawrinka","Juan Martin del Potro","Marin Cilic","Sam Querrey","Kevin Anderson","Jo-Wilfried Tsonga","Total")

#Look at yearly winnings
year.net.875 <- rbind(nadal.year.net.875,federer.year.net.875,djokovic.year.net.875,wawrinka.year.net.875,delpotro.year.net.875,cilic.year.net.875,queery.year.net.875,anderson.year.net.875,tsonga.year.net.875)
year.net.875 <- year.net.875 %>% group_by(year) %>% summarise(net=sum(net))

#0.9 odds
net.9 <- c(nadal.net.9,federer.net.9,djokovic.net.9,wawrinka.net.9,delpotro.net.9,cilic.net.9,queery.net.9,anderson.net.9,tsonga.net.9)
wins.9 <- c(nadal.wins.9,federer.wins.9,djokovic.wins.9,wawrinka.wins.9,delpotro.wins.9,cilic.wins.9,queery.wins.9,anderson.wins.9,tsonga.wins.9)
tries.9 <- c(nadal.tries.9,federer.tries.9,djokovic.tries.9,wawrinka.tries.9,delpotro.tries.9,cilic.tries.9,queery.tries.9,anderson.tries.9,tsonga.tries.9)

df.9 <- data.frame("Net"= net.9,"Wins"=wins.9,"Tries"=tries.9)
df.9["Total" ,] <- colSums(df.9)
df.9 <- mutate(df.9, "Win Percentage" = Wins/Tries)
row.names(df.9) <- c("Rafael Nadal","Roger Federer","Novak Djokovic","Stan Wawrinka","Juan Martin del Potro","Marin Cilic","Sam Querrey","Kevin Anderson","Jo-Wilfried Tsonga","Total")


#Look at yearly winnings
year.net.9 <- rbind(nadal.year.net.9,federer.year.net.9,djokovic.year.net.9,wawrinka.year.net.9,delpotro.year.net.9,cilic.year.net.9,queery.year.net.9,anderson.year.net.9,tsonga.year.net.9)
year.net.9 <- year.net.9 %>% group_by(year) %>% summarise(net=sum(net))
```


```{r}
#display information for simulation with minimum =0.85
df.85
year.net.85
#display information for simulation with minimum =0.875
df.875
year.net.875
#display information for simulation with minimum =0.9
df.9
year.net.9
```

```{r, echo=FALSE}
my.theme <- theme(axis.text.x=element_text(face="bold",color="cyan4", size=8), axis.text.y=element_text(face="bold",color="cyan4",size=8), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5), plot.subtitle = element_text(face="bold", color="violetred4", size=9, hjust=.5))

#plot for probability of 0.85
ggplot(data=df.85[1:(dim(df.85)[1]-1),],aes(x=rownames(df.85[1:(dim(df.85)[1]-1),]),y=Net))+geom_bar(stat="identity",fill = "orange3")+labs(title="Simulation of Betting on 2015, 2016, and 2017 Seasons",subtitle="when Wagering $10 when Logistic Regression Predicts Probability of Winning Greater than 0.85",x="Player",y="Net Winnings")+my.theme+ theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplot(data=year.net.85,aes(x=year,y=net))+geom_bar(stat="identity",fill = "orange3")+labs(title="Simulation of Betting on 2015, 2016, and 2017 Seasons",subtitle="when Wagering $10 when Logistic Regression Predicts Probability of Winning Greater than 0.85",x="Year",y="Net Winnings")+my.theme

#plots for probability of 0.875
ggplot(data=df.875[1:(dim(df.875)[1]-1),],aes(x=rownames(df.875[1:(dim(df.875)[1]-1),]),y=Net))+geom_bar(stat="identity",fill = "orange3")+labs(title="Simulation of Betting on 2015, 2016, and 2017 Seasons",subtitle="when Wagering $10 when Logistic Regression Predicts Probability of Winning Greater than 0.875",x="Player",y="Net Winnings")+my.theme+ theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplot(data=year.net.875,aes(x=year,y=net))+geom_bar(stat="identity",fill = "orange3")+labs(title="Simulation of Betting on 2015, 2016, and 2017 Seasons",subtitle="when Wagering $10 when Logistic Regression Predicts Probability of Winning Greater than 0.875",x="Year",y="Net Winnings")+my.theme

#plots for probability of 0.9
ggplot(data=df.9[1:(dim(df.9)[1]-1),],aes(x=rownames(df.9[1:(dim(df.9)[1]-1),]),y=Net))+geom_bar(stat="identity",fill = "orange3")+labs(title="Simulation of Betting on 2015, 2016, and 2017 Seasons",subtitle="when Wagering $10 when Logistic Regression Predicts Probability of Winning Greater than 0.9",x="Player",y="Net Winnings")+my.theme+ theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplot(data=year.net.9,aes(x=year,y=net))+geom_bar(stat="identity",fill = "orange3")+labs(title="Simulation of Betting on 2015, 2016, and 2017 Seasons",subtitle="when Wagering $10 when Logistic Regression Predicts Probability of Winning Greater than 0.9",x="Year",y="Net Winnings")+my.theme


```

With this betting strategy, a new loss occurred when setting the probability minimum probability of winning a match for matches we would bet on at 0.85 and 0.9. A profit of $31.3825 was won on on 423 bets when the minimum probability was set at 0.875. The model predicts the outcome of the matches fairly accurately as seen in the win percentages of the data frames above. This accurate prediction of the probability of winning does not translate to net winnings that are very high because you are most likely placing bets when the payoff for being correct is not that high. Therefore, even if your bet is wrong only a small percentage of the time as it is above, the loss of your wager for these few incorrect bets is close to the amount of gain you have from all of your correct predictions. This also shows that always betting on the favorite or those who are ranked higher will not work. There will sometimes be matches in which a heavily favored player loses, so betting on safe matches will not work effectively. Being risk averse and only betting on the safe matches does not translate to winnings.

