---
title: "Machine Learning Modeling"
author: "Will Renius"
date: "November 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries
```{r}
#install.packages("fastAdaboost")
#install.packages("randomForest")
#install.packages("e1071")
#detach("package:stats")
library(tidyverse)
library(lubridate)
library(fastAdaboost)
library(e1071)
library(randomForest)
library(scales)
library (ROCR)
```

#load Data
```{r}

load("data/favorite.data.RData")
men.data <- favorite.data
```

View Data Structure
```{r}
str(men.data)
```

## Impute Data
Imputing data replaces any null values. This is necessary for some models that will break when encountering nulls.
```{r}
#create new dataframe to impute
imputed.data <- men.data

#add favorite_won column used to predict winner
imputed.data$favorite_won <- as.factor(!imputed.data$odds.upset)

#Convert characters to factors
character.columns <- c("Location", "Tournament", "Series", "Surface", "Court","Round" )
for(col.name in character.columns){
  imputed.data[[col.name]]<- as.factor(imputed.data[[col.name]])
}

#List of odds books columns
favorite.columns <- c("B365F", "B.WF" , "CBF" , "EXF" , "LBF" , "GBF" , "IWF" , "PSF" , "SBF" , "SJF" , "UBF" )
underdog.columns <- c("B365U", "B.WU" , "CBU" , "EXU" , "LBU" , "GBU" , "IWU" , "PSU" , "SBU" , "SJU" , "UBU")

#replace all null value books with the average over all books for that match
for(col.name in favorite.columns){
  imputed.data[[col.name]][is.na(imputed.data[[col.name]])] <- imputed.data[["avgF"]][is.na(imputed.data[[col.name]])]
}
for(col.name in underdog.columns){
  imputed.data[[col.name]][is.na(imputed.data[[col.name]])] <- imputed.data[["avgU"]][is.na(imputed.data[[col.name]])]
}

#remove any remaining null values, and grab only the columns that we can use in our models
imputed.data <- imputed.data %>% na.omit() %>% select(-maxWinOdds, -minLoseOdds, -maxProfit, -rank.upset, -odds.upset, -Tournament, -Location)
#create a column for rank difference
imputed.data <- imputed.data %>% mutate(rank.diff = abs( URank - FRank))

#view structure of final imputed data
str(imputed.data)

```
Theme for plots
```{r}
# theme
my.theme <- theme(axis.text.x=element_text(face="bold",color="cyan4", size=10), axis.text.y=element_text(face="bold",color="cyan4",size=10), axis.title = element_text(face="bold",color="royalblue4",size=13), plot.title = element_text(face="bold",color="violetred4",size=15,hjust=.5), legend.title=element_blank(), plot.subtitle = element_text(face="bold", color="royalblue4", size=10, hjust=.5))

```


```{r}
#Function for plotting winnings over time
plot.winnings <- function(df, title){
  df.new.pred <- df %>% dplyr::filter(is.finite(winnings)) %>% arrange(date.new) %>% mutate(winnings = cumsum(winnings))
  plot.df <- ggplot(df.new.pred, aes(date.new, winnings))  + geom_line(color='orange3', size=1) + xlab("Date") + ylab("Winnings") + ggtitle(title)
  
  print(plot.df)
}
```

# Use Models to classify whether winner was the favorite or underdog
```{r}
#create lists to keep track of the final payout each year
rf.final.payouts = c()
svm.final.payouts =c()
reg.final.payouts = c()
boost.final.payouts = c()

#starting in 2011 because some data causes trouble before this year.
for(i in 2017:2017){
  #create date strings to use to limit training and testing datasets
  train.start.date <-  paste0(i-3, "-01-01", sep = '')
  train.end.date <- paste0(i-1, "-01-01", sep = '')
  test.start.date <- paste0(i, "-01-01", sep = '')
  test.end.date <- paste0(i+1, "-01-01", sep = '')
  
  #view year start of testing set
  print(test.start.date)
  
  #create training and testing dataset
  new.train <- imputed.data %>% dplyr::filter(date.new < as.Date(train.end.date), date.new >= as.Date(train.start.date)) 
  test.data <- imputed.data %>% dplyr::filter(date.new > as.Date(test.start.date), date.new < as.Date(test.end.date))
  
  #Make predictive models
  
  rf.model = randomForest(favorite_won~., data=new.train, ntree=20, proximity=T)

  #SVM
  svm.model <- svm(favorite_won~., data = new.train)
  
  #Log Reg
  reg.model <- glm(favorite_won ~.,family=binomial(link='logit'),data=new.train)
  
  #Addaboost
  boost.model <- adaboost(favorite_won ~., new.train, nIter = 20)
                   
  
  #Get boolean true false whether model predicts an upset or not
  rf.prediction = predict(rf.model, newdata=test.data)
  svm.prediction <- predict(svm.model, test.data)
  reg.prediction <- as.factor(predict(reg.model, test.data, type = "response") > .5)
  boost.prediction <- predict( boost.model,test.data)$class

  
  #Make data with prediction column for each match

  #random forest
  rf.pred <- test.data
  rf.pred$prediction <- rf.prediction

  #SVM
  svm.pred <- test.data
  svm.pred$prediction <- svm.prediction

  #Log Regression
  reg.pred <- test.data
  reg.pred$prediction <- reg.prediction

  
  #Addaboost
  boost.pred <- test.data
  boost.pred$prediction <- boost.prediction

  test.data$print_favorite_won = ifelse(test.data$favorite_won == "TRUE", "True: Favorite Won", "False: Underdog Won")
  print("reg")
  CM = table(test.data$print_favorite_won, reg.pred$prediction)
  accuracy = (sum(diag(CM)))/sum(CM)
  print(CM)
  print(paste("accuracy: ",accuracy ))

  
  print("svm")
  
  CM = table(test.data$print_favorite_won, svm.pred$prediction)
  accuracy = (sum(diag(CM)))/sum(CM)
  print(CM)
  print(accuracy)

  print("rf")
  CM = table(test.data$print_favorite_won, rf.pred$prediction)
  accuracy = (sum(diag(CM)))/sum(CM)
  print(CM)
  print(accuracy)

  
  print("boost")
  CM = table(test.data$print_favorite_won, boost.pred$prediction)
  accuracy = (sum(diag(CM)))/sum(CM)
  print(CM)
  print(paste("accuracy: ",accuracy ))
  print(CM)

  
  #make bets for each model 
  bet <- 10
  rf.bets <- rf.pred %>% mutate( winnings = case_when(prediction != favorite_won ~-bet,
                                             prediction == "TRUE" & favorite_won == prediction  ~ bet * (FMaxOdds -1),
                                             prediction == "FALSE" & favorite_won == prediction ~ bet * (UMaxOdds -1),
                                             TRUE ~ NA_real_))
  svm.bets <- svm.pred %>% mutate( winnings = case_when(prediction != favorite_won ~-bet,
                                             prediction == "TRUE" & favorite_won == prediction  ~ bet * (FMaxOdds -1),
                                             prediction == "FALSE" & favorite_won == prediction ~ bet * (UMaxOdds -1),
                                             TRUE ~ NA_real_))
  reg.bets <- reg.pred %>% mutate( winnings = case_when(prediction != favorite_won ~-bet,
                                             prediction == "TRUE" & favorite_won == prediction  ~ bet * (FMaxOdds -1),
                                             prediction == "FALSE" & favorite_won == prediction ~ bet * (UMaxOdds -1),
                                             TRUE ~ NA_real_))
  boost.bets <- boost.pred %>% mutate( winnings = case_when(prediction != favorite_won ~-bet,
                                             prediction == "TRUE" & favorite_won == prediction  ~ bet * (FMaxOdds -1),
                                             prediction == "FALSE" & favorite_won == prediction ~ bet * (UMaxOdds -1),
                                             TRUE ~ NA_real_))
  
  
  #Plot winnings over time for each model

  plot.winnings(rf.bets, paste(test.start.date, "Random Forest Winnings"))
  plot.winnings(svm.bets, paste(test.start.date, "Support Vector Machine Winnings"))
  plot.winnings(reg.bets, paste(test.start.date, "Logistic Regression Winnings"))
  plot.winnings(boost.bets, paste(test.start.date, "AdaBoost Winnings"))

  #append final payouts
  rf.final.payouts <-  c(rf.final.payouts, sum(rf.bets$winnings))
  svm.final.payouts <-  c(svm.final.payouts, sum(svm.bets$winnings))
  reg.final.payouts <-  c(reg.final.payouts, sum(reg.bets$winnings))
  boost.final.payouts <-  c(boost.final.payouts, sum(boost.bets$winnings))
  
  #make aggregate table
  agg.table <- rf.bets %>% summarise(test.start.date =test.start.date, method = "Random Forest",count = n(), net.profit = sum(winnings), successful.bets = sum(rf.bets$prediction == rf.bets$favorite_won))
  agg.table <- agg.table %>%
     rbind(svm.bets  %>% summarise(test.start.date =test.start.date, method = "SVM",count = n(), net.profit = sum(winnings), successful.bets = sum(svm.bets$prediction ==svm.bets$favorite_won))) %>%
    rbind(reg.bets  %>% summarise(test.start.date =test.start.date, method = "Log reg",count = n(), net.profit = sum(winnings), successful.bets = sum(reg.bets$prediction ==reg.bets$favorite_won))) %>%
  rbind(boost.bets  %>% summarise(test.start.date =test.start.date, method = "Boost",count = n(), net.profit = sum(winnings), successful.bets = sum(boost.bets$prediction ==boost.bets$favorite_won)))

  agg.table <- agg.table %>%mutate(success.rate = successful.bets / count, average.payout = net.profit / count)
  print(agg.table)
}
final.payouts <- data.frame(random.forest = rf.final.payouts, svm = svm.final.payouts, log.reg = reg.final.payouts, ada.boost = boost.final.payouts)

final.payouts
```
# Use Models to predict how much money you will win if you bet on favorite or underdog

First we need to change imputed data for our new prediction column
```{r}
#create new dataframe to impute
imputed.data <- men.data

#add favorite_won column used to predict winner
imputed.data$favorite_won <- as.factor(!imputed.data$odds.upset)

#Convert characters to factors
character.columns <- c("Location", "Tournament", "Series", "Surface", "Court","Round" )
for(col.name in character.columns){
  imputed.data[[col.name]]<- as.factor(imputed.data[[col.name]])
}

#List of odds books columns
favorite.columns <- c("B365F", "B.WF" , "CBF" , "EXF" , "LBF" , "GBF" , "IWF" , "PSF" , "SBF" , "SJF" , "UBF" )
underdog.columns <- c("B365U", "B.WU" , "CBU" , "EXU" , "LBU" , "GBU" , "IWU" , "PSU" , "SBU" , "SJU" , "UBU")

#replace all null value books with the average over all books for that match
for(col.name in favorite.columns){
  imputed.data[[col.name]][is.na(imputed.data[[col.name]])] <- imputed.data[["avgF"]][is.na(imputed.data[[col.name]])]
}
for(col.name in underdog.columns){
  imputed.data[[col.name]][is.na(imputed.data[[col.name]])] <- imputed.data[["avgU"]][is.na(imputed.data[[col.name]])]
}

#remove any remaining null values, and grab only the columns that we can use in our models
imputed.data <- imputed.data %>% na.omit() %>% select(-maxWinOdds, -minLoseOdds, -maxProfit, -rank.upset, -odds.upset, -Tournament, -Location)
#create a column for rank difference
imputed.data <- imputed.data %>% mutate(rank.diff = abs( URank - FRank))

#note if you change this vaue here and not above, it will be difficult to compare between methods
bet <- 10

#make payout columns and select columns ou need
imputed.data <- imputed.data %>% na.omit() %>% mutate(payout.favorite = case_when(favorite_won == "TRUE"  ~ bet * ( FMaxOdds - 1), favorite_won =="FALSE" ~ -bet, TRUE ~ NA_real_), payout.underdog = case_when(favorite_won == "FALSE"  ~ bet * ( UMaxOdds - 1), favorite_won =="TRUE" ~ -bet, TRUE ~ NA_real_)) 

# choose payout column to be the underdog(can switch to favorite in mutate)
imputed.data <- imputed.data %>% mutate(payout = payout.underdog) %>% select(-payout.favorite, -payout.underdog, -favorite_won)

#view new data
str(imputed.data)
```

```{r}
rf.final.payouts = c()
svm.final.payouts =c()
reg.final.payouts = c()
for(i in 2017:2017){
  #create date strings to use to limit training and testing datasets
  train.start.date <-  paste0(i-3, "-01-01", sep = '')
  train.end.date <- paste0(i-1, "-01-01", sep = '')
  test.start.date <- paste0(i, "-01-01", sep = '')
  test.end.date <- paste0(i+1, "-01-01", sep = '')
  
  print(test.start.date)
  
  #create training and testing data
  new.train <- imputed.data %>% dplyr::filter(date.new < as.Date(train.end.date), date.new >= as.Date(train.start.date)) 
  test.data <- imputed.data %>% dplyr::filter(date.new > as.Date(test.start.date), date.new < as.Date(test.end.date))

  #Make predictive models
  
  rf.model <-  randomForest(payout~., data=new.train, ntree=20, proximity=T)
  
  #SVM
  svm.model <- svm(payout~., data = new.train)
  
  #Log Reg
  reg.model <- glm(payout ~.,data=new.train)
  
  
  
  
  rf.prediction <- predict(rf.model, newdata=test.data)
  svm.prediction <- predict(svm.model, test.data)
  reg.prediction <- predict(reg.model, test.data)


  #Make data with prediction column
  
  #random forest
  rf.pred <- test.data
  rf.pred$prediction <- rf.prediction
  
  #SVM
  svm.pred <- test.data
  svm.pred$prediction <- svm.prediction
  
  #Log Regression
  reg.pred <- test.data
  reg.pred$prediction <- reg.prediction
  
  # print(paste("rf R.squared: ", cor(test.data$payout,rf.prediction ) ^ 2))
  # print(paste("SVM R.squared: ", cor(test.data$payout,svm.prediction ) ^ 2))
  # print(paste("Logistic R.squared: ", cor(test.data$payout,reg.prediction ) ^ 2))
  
  #View the predictions and results in order of highest expected return
  rf.pred.group <- rf.pred %>% group_by(prediction, payout) %>% summarize(count = n()) %>% arrange(desc(prediction))
  svm.pred.group <- svm.pred %>% group_by(prediction, payout) %>% summarize(count = n()) %>% arrange(desc(prediction))
  reg.pred.group <- reg.pred %>% group_by(prediction, payout) %>% summarize(count = n()) %>% arrange(desc(prediction))

  # print(rf.pred.group)
  # print(svm.pred.group)
  # print(reg.pred.group)

  
  #make bets, here you can filter for when predicted winnings are greate than some threshold.
  rf.bets <- rf.pred %>% mutate( winnings = payout) %>% dplyr::filter(prediction > 2)
  svm.bets <- svm.pred %>% mutate( winnings = payout) %>% dplyr::filter(prediction > 2)
  reg.bets <- reg.pred %>% mutate( winnings = payout) %>% dplyr::filter(prediction > 2)

  #plot winnings
  plot.winnings(rf.bets, paste(test.start.date, "Random Forest Payout"))
  plot.winnings(svm.bets, paste(test.start.date, "Support Vector Machine Payout"))
  plot.winnings(reg.bets, paste(test.start.date, "Logistic Regression payout"))
 
  
  #append final payouts
  rf.final.payouts <-  c(rf.final.payouts, sum(rf.bets$winnings))
  svm.final.payouts <-  c(svm.final.payouts, sum(svm.bets$winnings))
  reg.final.payouts <-  c(reg.final.payouts, sum(reg.bets$winnings))
  
  #create aggregate table
  agg.table <- rf.bets %>% summarise(test.start.date =test.start.date, method = "Random Forest",count = n(), net.profit = sum(winnings), successful.bets = sum(rf.bets$winnings > 0))
  agg.table <- agg.table %>%
     rbind(svm.bets  %>% summarise(test.start.date =test.start.date, method = "SVM",count = n(), net.profit = sum(winnings), successful.bets = sum(svm.bets$winnings > 0))) %>%
    rbind(reg.bets  %>% summarise(test.start.date =test.start.date, method = "Log reg",count = n(), net.profit = sum(winnings), successful.bets = sum(reg.bets$winnings > 0))) 
  
  print(agg.table)
}
final.payouts <- data.frame(random.forest = rf.final.payouts, svm = svm.final.payouts, log.reg = reg.final.payouts)
final.payouts
```

 






