---
title: "Random Forest"
author: "Will Renius"
date: "November 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries
```{r}
#detach("package:stats")
library(tidyverse)
library(lubridate)
```


```{r}
load("data/clean.men.data.RData")
men.data <- final.data
```

```{r}
str(men.data)
```


```{r}
men.data %>% group_by(upset) %>% summarize(count = n())

```


```{r}
train.data <- men.data %>% dplyr::filter(date.new < as.Date("2017-01-01")) %>% select(-date.new)
test.data <- men.data %>% dplyr::filter(date.new > as.Date("2017-01-01")) %>% select(-date.new)
#, date.new < as.Date("2018-01-01")
```


```{r}
require(randomForest)
imputed.data <- men.data
imputed.data$favorite_won <- as.factor(imputed.data$favorite_won)
favorite.columns <- c("B365F", "B.WF" , "CBF" , "EXF" , "LBF" , "GBF" , "IWF" , "PSF" , "SBF" , "SJF" , "UBF" )
underdog.columns <- c("B365U", "B.WU" , "CBU" , "EXU" , "LBU" , "GBU" , "IWU" , "PSU" , "SBU" , "SJU" , "UBU")

for(col.name in favorite.columns){
  imputed.data[[col.name]][is.na(imputed.data[[col.name]])] <- imputed.data[["avgF"]][is.na(imputed.data[[col.name]])]
}
for(col.name in underdog.columns){
  imputed.data[[col.name]][is.na(imputed.data[[col.name]])] <- imputed.data[["avgU"]][is.na(imputed.data[[col.name]])]
}

imputed.data <- imputed.data%>% select(-maxWinOdds, -minLoseOdds, -maxProfit, -upset)
str(imputed.data)

```


```{r}
new.train <- imputed.data %>% dplyr::filter(date.new < as.Date("2017-01-01"), date.new >= as.Date("2015-01-01")) %>% select(-date.new)
# new.train <- sample_frac(train.imputed, .1, replace = F)
iris_rf = randomForest(favorite_won~., data=new.train, ntree=200, proximity=T)
table(predict(iris_rf), new.train$favorite_won)

```

```{r}
testData <- imputed.data %>% dplyr::filter(date.new > as.Date("2010-01-01"), date.new < as.Date("2011-01-01")) %>% select(-date.new)


iris_rf
plot(iris_rf)
importance(iris_rf)
irisPred = predict(iris_rf, newdata=testData)
table(irisPred, testData$favorite_won)
#plot(margin(iris_rf, testData$favorite_won))
test.pred <- testData
test.pred$prediction <- irisPred
str(test.pred)
head(test.pred, 1)
```

```{r}
test.pred <- test.pred %>% mutate(maxProfit = case_when(favorite_won == "TRUE"  ~ 100 * ( FMaxOdds - 1), favorite_won =="FALSE" ~ 100 * (UMaxOdds - 1)  , TRUE ~ NA_real_))

test.pred <- test.pred %>% mutate( winnings = case_when( prediction == "FALSE" & avgU > 3 & favorite_won == prediction  ~ maxProfit * 1,  prediction == "FALSE" & avgU > 3 & favorite_won != prediction ~ -100 * 1, TRUE ~ NA_real_))

sum(test.pred$winnings, na.rm = T)
```

```{r}
#test.pred %>% group_by(maxProfit) %>% summarise(count = n())
```

```{r}
test.pred %>% group_by(prediction, favorite_won) %>% summarize(avgF = mean(avgF), avgU = mean(avgU), n = n(), bets = sum(!is.na(winnings)), earnings = sum(winnings, na.rm = T))
```

```{r}
naive <- test.pred %>% mutate( winnings = case_when(upset == F & countWinOdds > 0 & countLoseOdds > 0 ~ maxProfit * 1, upset != F & countWinOdds > 0 & countLoseOdds > 0 ~ -1 * 1, TRUE ~ NA_real_))
sum(naive$winnings, na.rm = T)
naive %>% group_by(upset, prediction) %>% summarise(sum(winnings, na.rm = TRUE))


```


```{r}
CM = table(irisPred, testData$favorite_won)
accuracy = (sum(diag(CM)))/sum(CM)
print(CM)
print(accuracy)
```




Now We have ability to predict the winner with hi

```{r}
# ```{r}
# print(forest)
# ```
# ```{r}
# newPred = predict(forest, newdata=test.imputed)
# table(newPred, test.imputed$upset)
# ```
# ```{r}
# plot(margin(forest, test.imputed$upset))
# 
# ```
# ```{r}
# CM = table(irisPred, testData$Species)
# accuracy = (sum(diag(CM)))/sum(CM)
# print
# ```
# 
# 
# 
# ```{r}
# importance(forest)
# ```
# 
# ```{r}
# #test.imputed <-  rfImpute(x = test.data, y = test.data$upset)
# # t1 <- test.data
# # t1[is.na(t1)] <- NaN
# # head(t1, 1)
# # test.imputed <-  rfImpute(upset ~ .,t1 %>% select(upset, avgW, avgL))
# # train.imputed <- rfImpute(upset ~ .,train.data)
# 
# # Boston.rf=randomForest( x = select(train.imputed, -upset) , y = train.imputed$upset, xtest = select(test.imputed,-upset), ytest = test.imputed$upset)
# # plot(Boston.rf)
# ```
# ```{r}
# 
# ```
# 
# ```{r}
# #rfImpute
# ```
# 
# ```{r}
# 
# ```
# 
# 
# 
# 
# ```{r}
# save(men.data, file = "men.data.RData")
# 
# ```
```


