---
title: "Untitled"
author: "Yifei Yan"
date: "2017Äê11ÔÂ26ÈÕ"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
library(dplyr)
library(ggplot2)
library(tidyr)

```


```{r}
load("favorite.data.RData")
```

```{r}
men.data <- favorite.data
```

```{r}
str(men.data)
```

### upset
```{r}
odds.upset_prop <- favorite.data %>% group_by(odds.upset) %>% summarize(count = n(), proportion = n()/dim(final.data)[1])
as.data.frame(odds.upset_prop)
```

```{r}
ggplot(data = data.frame(odds.upset_prop), mapping = aes(x = odds.upset, y= proportion ))+geom_bar(stat = "identity")+
  scale_y_continuous(labels = scales :: percent)+
  labs(x = "UPSET",y=NULL, title = "Proportions in Rank Upset ")


```
### odds.upset vs Surface
```{r}
Surface_prop <- favorite.data %>% group_by(Surface) %>% summarize(count = n(), prop.upset = mean(as.logical(odds.upset)))%>% mutate(prop.fav = 1-prop.upset)
as.data.frame(Surface_prop)

```


```{r}
ggplot(data = data.frame(Surface_prop),aes(x = Surface, y= prop.upset, fill =Surface))+geom_bar(stat = "identity", position="dodge")+
  scale_y_continuous(labels = scales :: percent)+labs(x = "Surface",y="percentage", title = "Proportions in Odds.Upset by Surface")
ggplot(data = data.frame(Surface_prop),aes(x = Surface, y= prop.fav, fill =Surface))+geom_bar(stat = "identity", position="dodge")+
  scale_y_continuous(labels = scales :: percent)+labs(x = "Surface",y="percentage", title = "Proportions in Odds.Fav by Surface")
```
### odds.upset vs Round
```{r}
Round_prop <- favorite.data %>% group_by(Round) %>% summarize(count = n(), prop.upset = mean(as.logical(odds.upset)))%>% mutate(prop.fav = 1-prop.upset)
as.data.frame(Round_prop)
```
```{r}
ggplot(data = data.frame(Round_prop),aes(x = Round, y= prop.upset, fill =Round))+geom_bar(stat = "identity", position="dodge")+
  scale_y_continuous(labels = scales :: percent)+labs(x = "Surface",y="percentage", title = "Proportions in Odds.Upset by Round")
ggplot(data = data.frame(Round_prop),aes(x = Round, y= prop.fav, fill =Round))+geom_bar(stat = "identity", position="dodge")+
  scale_y_continuous(labels = scales :: percent)+labs(x = "Surface",y="percentage", title = "Proportions in Odds.Fav by Round")
```
### odds.upset vs Surface & Round
```{r}
SurfaceRound_prop <- favorite.data %>% group_by(Surface,Round) %>% summarize(count = n(), prop.upset = mean(as.logical(odds.upset)))%>% mutate(prop.fav = 1-prop.upset)
as.data.frame(SurfaceRound_prop)
```


```{r}
ggplot(data = data.frame(SurfaceRound_prop), mapping = aes(x = Surface, y= prop.upset , fill = Surface))+geom_bar(stat = "identity", position="dodge")+
  scale_y_continuous(labels = scales :: percent)+labs(x = "Surface",y="percentage", title = "Proportions in Odds.Upset by Surface")+ facet_grid(~Round)+theme(text = element_text(size=10), axis.text.x = element_text(angle=45, hjust=1)) 
ggplot(data = data.frame(SurfaceRound_prop), mapping = aes(x = Surface, y= prop.fav , fill = Surface))+geom_bar(stat = "identity", position="dodge")+
  scale_y_continuous(labels = scales :: percent)+labs(x = "Surface",y="percentage", title = "Proportions in Odds.Upset by Surface")+ facet_grid(~Round)+theme(text = element_text(size=10), axis.text.x = element_text(angle=45, hjust=1)) 
```


###Upset by Series
The Grandslam tourments, also called major, are the four most important annual tennis events. They offer most prize pool, ranking points and public attention.
According to the graph below, we can clearly see that Grand Slam has lowest upset rates. So keep ketting on odds favorite should be a solid strategy on Grand Slam Series. 
```{r}
Series <- favorite.data %>% group_by(Series) %>% summarize(count = n(), prop.upset = mean(as.logical(odds.upset)))%>% mutate(prop.fav = 1-prop.upset)
as.data.frame(Series)

ggplot(data = data.frame(Series), mapping = aes(x = Series, y= prop.upset , fill = Series))+geom_bar(stat = "identity", position="dodge",colours = "Series")+
scale_y_continuous(labels = scales :: percent)+labs(x = "upset",y="percentage", title = "Proportions in odds.upset by Series") + theme(text = element_text(size=15), axis.text.x = element_text(angle=45, hjust=1)) 

ggplot(data = data.frame(Series), mapping = aes(x = Series, y= prop.fav , fill = Series))+geom_bar(stat = "identity", position="dodge",colours = "Series")+
scale_y_continuous(labels = scales :: percent)+labs(x = "Series",y="percentage", title = "Proportions in odds.fav by Series") + theme(text = element_text(size=15), axis.text.x = element_text(angle=45, hjust=1)) 
```


```{r}

Grandslam.upset <- favorite.data %>% group_by(Series,Round) %>% summarize(count = n(), prop.upset = mean(as.logical(odds.upset)))%>% mutate(prop.fav = 1-prop.upset)
as.data.frame(Grandslam.upset)
filter(Grandslam.upset, Series == "Grand Slam")

ggplot(data = data.frame(Grandslam.upset), mapping = aes(x = Round, y= prop.upset,fill =Round ))+geom_bar(stat = "identity", position="dodge")+
  scale_y_continuous(labels = scales :: percent)+
  labs(x = "UPSET",y="Percentage", title = "Proportions in UPSET by Grand Slam & Round")+theme(text = element_text(size=10), axis.text.x = element_text(angle=45, hjust=1)) 
```









